### 2013-11-07 src2/thesis.tex
update

This thesis applies embedded language methodology to data-parallel programming of 
massively parallel hardware. Embedded languages are languages that are implemented 
as a library for some existing host language. In this case, the functional programming 
language Haskell is used as host language.

@@

Before CUDA, programmers still tried (very ingeniously) to make use of GPUs for 
non-graphics applications via the graphics API. The computation needed to be expressed 
in graphics terms and the result computed through an act of rendering. Of course, since the 
available abstractions targetted at the graphics domain, this added extra
burden  on the programmer. CUDA was NVIDIA's first response to this problem. With CUDA 
there is a dialect of C, extended with GPU specific functionality.

Having a language specifically made for general purpose programming on the GPU, such as 
CUDA, is a big step up from earlier. But CUDA still offers a very low-level interface to 
programming the GPU, which is preferred in a situation where the programmer needs to 
think about hardware specific details to get the best performance. This CUDA provides, 
but at the price that it is hard to build upon, abstract and reuse prior programming 
effort. This is where we are trying to go one step further with Obsidian.

@@

@@

\begin{small} 
\begin{Verbatim}[samepage=true]
kernel<<<1,32,0>>>(d_i1,d_i2,d_r);
\end{Verbatim}
\end{small}

@@

\begin{small}
\begin{Verbatim}[samepage=true]

\end{Verbatim]
\end{small}

@@

% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
%
% GPU Programming 
%
% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
\section{GPUs and CUDA}

In 2006 NVIDIA released their first Compute Unified Device Architecture (CUDA) 
GPU. With CUDA, NVIDIA also made it a lot easier for programmers to use GPUs 
for general purpose programming. A dialect of C, called CUDA C, was released specifically 
to simplify implementation of general purpose algorithms on GPUs.

A CUDA GPU is built around a single kind of processor (as opposed to the different 
kinds of processors found in earlier GPUs). The processors in the GPU (called MPs, 
MultiProcessors) all contain a number of function units called SPs (Streaming Processors). 
Each MP also contains local memory, called shared memory since it can be accessed 
by all of the SPs in that MP. The number of MPs varies over the available GPUs; cheaper 
GPUs have as few as one MP, and as you go up in price the number of MPs increases.

Each MP of the GPU can manage a large number of threads; on today's GPUs up to 
2048 threads can run on a single MP. The GPU schedules threads in groups of 32, called 
{\em Warps}, that are executed in lock-step (SIMD style execution). Threads are 
also divided into {\em Blocks}; the threads within a block can communicate using the 
shared memory. The maximum block size is currently 1024 threads. Since 
the number of threads in a block can be higher than the warp size, a barrier 
synchronisation mechanism exists to ensure that threads within a block 
can communicate safely via the shared memory.

Papers B, C, E and F (sections~\ref{sec:paperB},~\ref{sec:paperC},~\ref{sec:paperE},
~\ref{sec:paperF}) contain more information about GPUs, and the assumptions we make 
about them in relation to Obsidian.

@@

  // Launch the kernel
  kernel<<<1,32,0>>>(d_i1,d_i2,d_r);

>>>>

This thesis applies embedded language methodology to data-parallel programming of 
massively parallel hardware. An Embedded language is a language that is implemented 
as a library for some existing host language. In this case, the functional programming 
language, Haskell is used as host language.

@@

Before CUDA, programmers still tried (very ingeniously) to make use of GPUs for 
non-graphics applications via the graphics API. The desired computation was expressed 
in graphics terms and the result computed through an act of rendering. Of course, since the 
available abstractions targeted the graphics domain, this added extra
burden on the programmer. CUDA was NVIDIA's first response to this problem. With CUDA 
there is a dialect of C, extended with a parallel programming model suitable for GPUs.

Having a language specifically made for general purpose programming on the GPU, such as 
CUDA, is a big step up from earlier. But CUDA still offers a very low-level interface to 
programming the GPU, which is preferred in a situation where the programmer needs to 
think about hardware specific details to get the best performance. CUDA provides this, 
but at the price that it is hard to build upon, abstract and reuse prior programming 
effort. This is where we are trying to go one step further with Obsidian.

@@

\begin{small}
\begin{Verbatim}[samepage=true]
__global__ void vecAdd(int32_t *i1, int32_t *i2, int32_t *r) {

@@

\begin{small} 
\begin{Verbatim}[samepage=true]
vecAdd<<<1,32,0>>>(d_i1,d_i2,d_r);
\end{Verbatim}
\end{small}

@@

\begin{small}
\begin{Verbatim}[samepage=true]
vecAdd :: Num a => SPull a -> SPull a -> SPull a
vecAdd = zipWith (+)
\end{Verbatim}
\end{small}

\noindent The {\tt vecAdd} program takes two {\tt SPull} input arrays and gives one 
as result. The Obsidian {\tt vecAdd} is more general than the CUDA version. From it 
we can generate CUDA kernels for any type that supports {\tt (+)}, is in the {\tt Num} class.

There are some important differences between CUDA and Obsidian programs. One is that with Obsidian 
we generate kernels for fixed size local computations. This is to avoid doing out-of-bounds check 
which would mean conditionals in the code and conditionals are problematic from a performance 
point of view. Another difference is that in Obsidian the programmer specifies the entire 
computation, including how it is mapped over the blocks of the GPU. This is done implicitly 
in CUDA, based on the indexing arithmetic used in the kernel and the number of blocks and 
threads provided at kernel launch. This means that the Obsidian programmer needs to 
write a program that performs this mapping:

\begin{small} 
\begin{Verbatim}[samepage=true]
vecAddG :: Num a => DPull a -> DPull a -> DPush Grid a
vecAddG i1 i2  =
  pConcat $
  fmap push $
  zipWith vecAdd
          (splitUp 32 i1)
          (splitUp 32 i2)  
\end{Verbatim}
\end{small}

@@

% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
%
% GPU Programming 
%
% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------

@@

  // Launch the kernel
  vecAdd<<<1,32,0>>>(d_i1,d_i2,d_r);
